---
title: Starting with Data
teaching: 70
exercises: 2
source: Rmd
editor_options: 
  markdown: 
    wrap: 72
---

::: instructor
The main goals for this lessons are:

-   To make sure that learners understand the structure of a dataframe
-   To expose learners to factors. Their behavior is not necessarily
    intuitive, and so it is important that they are guided through it
    the first time they are exposed to it. The content of the lesson
    should be enough for learners to avoid common mistakes with them.
:::

::: objectives
-   Load external data from a `.csv` file into a data frame.
-   Explore the structure and content of data.frames
-   Understand how `R` assigns values to objects
-   Understand vector types and missing data
-   Describe the difference between a factor and a string.
-   Create and convert factors
-   Examine and change date formats.
:::

::: questions
-   How does `R` store data?
-   What is a data.frame?
-   How can I read a complete `.csv` file into `R`?
-   How can I get basic summary information about my dataset?
-   How can I change the way `R` treats strings in my dataset?
-   Why would I want strings to be treated differently?
-   How are dates represented in `R` and how can I change the format?
:::

## Acknowledgement

This workshop was adapted using material from the Data Carpentry lessons
[`R for Social Scientists`](https://datacarpentry.github.io/r-socialsci/index.html),
specifically [`lesson 02-starting-with-data`](https://datacarpentry.github.io/r-socialsci/02-starting-with-data.html),
and [`R for Ecologists`](https://datacarpentry.github.io/R-ecology-lesson/index.html),
specifically
[`how-r-thinks-about-data`](https://datacarpentry.github.io/R-ecology-lesson/how-r-thinks-about-data.html).

## Other Materials

[See Workshop 3 Slides
here](https://irimmn.sharepoint.com/:p:/s/IRIMRWorkshops/IQARukopMq9ITq8gmPXMml9zAYqZU3vDsT6V57tMfkiLR9I?e=jJuvIK)

[See Workshop 3 recording
here](https://irimmn.sharepoint.com/:v:/s/IRIMRWorkshops/IQCOPuF89NgRT725GgpozJKhARghJz3XspT9d2XQTv5w0Qs?e=63Ozha)

## Set up

Start by opening up your `RStudio project` that you created in a [previous
workshop](https://kathrynnapier.github.io/irim-r-workshops/introduction-r-rstudio.html#getting-set-up-in-rstudio)
(called `intro_r`). Open a new `R Notebook`: `Click File -> New File -> R`
Notebook. Save your `R Notebook` with a filename that makes sense, such as
`starting_with_data.Rmd`, in the `scripts` folder.

When you open a new `R Notebook`, some explanatory text is provided. This
can be deleted so you can enter your own text and code.

## What are data frames?

Data frames are the *de facto* data structure for tabular data in `R`,
and what we use for data processing, statistics, and plotting.

A data frame is the representation of data in the format of a table
where the columns are vectors that all have the same length. Data frames
are analogous to the more familiar spreadsheet in programs such as
`Excel`, with one key difference. Because columns are vectors, each column
must contain a single type of data (e.g., characters, integers,
factors). For example, here is a figure depicting a data frame
comprising a numeric, a character, and a logical vector.

![](fig/data-frame.svg){alt="A 3 by 3 data frame with columns showing numeric, character and logical values."}

Data frames can be created by hand, but most commonly they are generated
by the functions `read_csv()` or `read_table()`; in other words, when
importing spreadsheets from your hard drive (or the web). We will now
demonstrate how to import tabular data using `read_csv()`.

## Presentation of the SAFI Data

`SAFI` (Studying African Farmer-Led Irrigation) is a study looking at
farming and irrigation methods in Tanzania and Mozambique. The survey
data was collected through interviews conducted between November 2016
and June 2017. For this lesson, we will be using a subset of the
available data. For information about the orginal dataset, see the
[dataset
description](https://datacarpentry.org/socialsci-workshop/index.html#data).

We will be using a subset of the dataset that has been provided
(`data/raw/SAFI_clean.csv`). In this dataset, the missing data is
encoded as `NULL`, each row holds information for a single interview
respondent, and the columns represent:

| column_name | description |
|-------------------------|-----------------------------------------------|
| key_id | Added to provide a unique Id for each observation. (The InstanceID field does this as well but it is not as convenient to use) |
| village | Village name |
| interview_date | Date of interview |
| no_membrs | How many members in the household? |
| years_liv | How many years have you been living in this village or neighboring village? |
| respondent_wall_type | What type of walls does their house have (from list) |
| rooms | How many rooms in the main house are used for sleeping? |
| memb_assoc | Are you a member of an irrigation association? |
| affect_conflicts | Have you been affected by conflicts with other irrigators in the area? |
| liv_count | Number of livestock owned. |
| items_owned | Which of the following items are owned by the household? (list) |
| no_meals | How many meals do people in your household normally eat in a day? |
| months_lack_food | Indicate which months, In the last 12 months have you faced a situation when you did not have enough food to feed the household? |
| instanceID | Unique identifier for the form data submission |

### Download the data

If you did not previously downloaded the `SAFI_clean.csv` dataset in the
[previous workshop](https://kathrynnapier.github.io/irim-r-workshops/introduction-r-packages-markdown.html#download-data),
please follow the instructions below to download it. If you already have
the file in your `data/raw/` folder, jump to the **Importing data**
section.

We will be using a dataset called `SAFI_clean.csv`. The direct download
link for this file is:
<https://github.com/datacarpentry/r-socialsci/blob/main/episodes/data/SAFI_clean.csv>.
This data is a slightly cleaned up version of the SAFI Survey Results
available on
[`figshare`](https://figshare.com/articles/dataset/SAFI_Survey_Results/6262019).

First, we need to create a new folder called `data` to store this
dataset. Go to the Files pane, and create a new folder named `data`, and
two subfolders called `cleaned` and `raw`.

```         
intro_r
│
└── scripts
│
└── data
│    └── cleaned
│    └── raw
│
└─── images
│
└─── documents
```

You can either download the `SAFI_clean.csv` dataset used for this
workshop from the `GitHub` link or with `R`. You can download the file from
this [`GitHub` link](https://github.com/datacarpentry/r-socialsci/blob/main/episodes/data/SAFI_clean.csv)
and save it as `SAFI_clean.csv` in the `data/raw` directory you just
created. Or you can do this directly from `R` by copying and pasting this
in your console:

`download.file(   "https://raw.githubusercontent.com/datacarpentry/r-socialsci/main/episodes/data/SAFI_clean.csv",   "data/raw/SAFI_clean.csv", mode = "wb"   )`

## Importing data

You are going to load the data in `R`'s memory using the function
`read_csv()` from the **`readr`** package, which is part of the
**`tidyverse`**; learn more about the **`tidyverse`** collection of
packages [here](https://www.tidyverse.org/). **`readr`** gets installed
as part as the **`tidyverse`** installation. When you load the
**`tidyverse`** (`library(tidyverse)`), the core packages (the packages
used in most data analyses) get loaded, including **`readr`**.

Before proceeding, however, this is a good opportunity to talk about
conflicts. Certain packages we load can end up introducing function
names that are already in use by pre-loaded `R` packages. For instance,
when we load the tidyverse package below, we will introduce two
conflicting functions: `filter()` and `lag()`. This happens because
`filter` and `lag` are already functions used by the stats package
(already pre-loaded in `R`). What will happen now is that if we, for
example, call the `filter()` function, `R` will use the `dplyr::filter()`
version and not the `stats::filter()` one. This happens because, if
conflicted, by default `R` uses the function from the most recently loaded
package. Conflicted functions may cause you some trouble in the future,
so it is important that we are aware of them so that we can properly
handle them, if we want.

To do so, we just need the following functions from the conflicted
package:

-   `conflicted::conflict_scout()`: Shows us any conflicted functions.\
-   `conflict_prefer("function", "package_prefered")`: Allows us to
    choose the default function we want from now on.

It is also important to know that we can, at any time, just call the
function directly from the package we want, such as `stats::filter()`.

Even with the use of an `RStudio` project, it can be difficult to learn
how to specify paths to file locations. Enter the **`here`** package! The
here package creates paths relative to the top-level directory (your
RStudio project). These relative paths work *regardless* of where the
associated source file lives inside your project, like analysis projects
with data and reports in different subdirectories. This is an important
contrast to using `setwd()`, which depends on the way you order your
files on your computer.

Before we can use the `read_csv()` and `here()` functions, we need to
load the `tidyverse` and `here` packages.

Add a new code chunk in your notebook, load the `tidyverse` and `here`
packages, and read in the `SAFI` dataset. We'll **assign** the dataset to
an object called `interviews`.

If you recall, the missing data is encoded as `NULL` in the dataset.
We'll tell this to the `read_csv()` function, so `R` will automatically
convert all the `NULL` entries in the dataset into `NA`.

```{r, eval=TRUE, message=FALSE, purl=FALSE}
library(tidyverse)
library(here)

interviews <- read_csv(
  here("data", "raw", "SAFI_clean.csv"), 
  na = "NULL")
```

In the above code, we notice the `here()` function takes folder and file
names as inputs (e.g., `"data"`, `"SAFI_clean.csv"`), each enclosed in
quotations (`""`) and separated by a comma. The `here()` will accept as
many names as are necessary to navigate to a particular file (e.g.,
`here("data", "raw", "SAFI_clean.csv)`).

The `here()` function can accept the folder and file names in an
alternate format, using a slash ("/") rather than commas to separate the
names. The two methods are equivalent, so that
`here("data", "raw", "SAFI_clean.csv")` and `here("data/raw/SAFI_clean.csv")`
produce the same result. (The slash is used on all operating systems;
backslashes are not used.)

## Assigning objects

In `R`, we can **assign** inputs to a named **object**. We do this using
the **assignment arrow** `<-`, <kbd>Alt+-</kbd> (`Windows`) or
<kbd>Option+-</kbd> (`Mac`).What we are doing here is taking the result of
the code on the right side of the arrow (reading in the csv file), and
assigning it to an object whose name is on the left side of the arrow
(`interviews`).

You may notice that the contents of the interviews data frame do not
display below the code cell. This is because assignments (`<-`) don't
display anything. If we want to check that our data has been loaded, we
can see the contents of the data frame by typing its name: `interviews`
into a new code chunk.

```{r, results="hold", purl=FALSE}
interviews
## Try also
## view(interviews)
## head(interviews)
```

## Exploring data frames

When working with the output of a new function, it's often a good idea
to check the `class()`:

```{r class-tibble}
class(interviews)
```

Whoa! What is this thing? It has multiple classes`spec_tbl_df`,
`tbl_df`, `tbl`, and `data.frame`? Well, it's called a `tibble`, and it
is the `tidyverse` version of a data.frame. It *is* a data.frame, but
with some added perks. It prints out a little more nicely, it highlights
`NA` values and negative values in red, and it will generally
communicate with you more (in terms of warnings and errors, which is a
good thing).

As a `tibble`, the type of data included in each column is listed in an
abbreviated fashion below the column names. For instance, here `key_ID`
is a column of floating point numbers (abbreviated `<dbl>` for the word
'double'), `respondent_wall_type` is a column of characters ( `<chr>`) and the`interview_date` is a column in the "date and time" format (`<dttm>`).

::: callout
**`tidyverse` vs. `base R`**

As we begin to delve more deeply into the `tidyverse`, we should briefly
pause to mention some of the reasons for focusing on the `tidyverse` set
of tools. In R, there are often many ways to get a job done, and there
are other approaches that can accomplish tasks similar to the
`tidyverse`.

The phrase **`base R`** is used to refer to approaches that utilize
functions contained in `R`'s default packages. We will use some `base R`
functions, such as `str()`, `head()`, and `nrow()`, and we will be using
more scattered throughout this workshop. However, there are some key
`base R` approaches we will not be teaching. These include square bracket
subsetting. You may come across code written by other people that looks
like `interviews[1:10, 2]`, which is a `base R` command. If you're
interested in learning more about these approaches, you can check out
other Carpentries lessons like the [`Software Carpentry Programming with R`](https://swcarpentry.github.io/r-novice-inflammation/) lesson.

We choose to teach the `tidyverse` set of packages because they share a
similar syntax and philosophy, making them consistent and producing
highly readable code. They are also very flexible and powerful, with a
growing number of packages designed according to similar principles and
to work well with the rest of the packages. The `tidyverse` packages
tend to have very clear documentation and wide array of learning
materials that tend to be written with novice users in mind. Finally,
the `tidyverse` has only continued to grow, and has strong support from
`RStudio`, which implies that these approaches will be relevant into the
future.
:::

::: callout
## Note

`read_csv()` assumes that fields are delimited by commas. However, in
several countries, the comma is used as a decimal separator and the
semicolon (;) is used as a field delimiter. If you want to read in this
type of files in `R`, you can use the `read_csv2` function. It behaves
exactly like `read_csv` but uses different parameters for the decimal
and the field separators. If you are working with another format, they
can be both specified by the user. Check out the help for `read_csv()`
by typing `?read_csv` to learn more. There is also the `read_tsv()` for
tab-separated data files, and `read_delim()` allows you to specify more
details about the structure of your file.
:::

## Inspecting data frames

When calling a `tbl_df` object (like `interviews`), there is already a
lot of information about our data frame being displayed such as the
number of rows, the number of columns, the names of the columns, and as
we just saw the class of data stored in each column. However, there are
functions to extract this information from data frames. Here is a
non-exhaustive list of some of these functions. Let's try them out!

Size:

-   `dim(interviews)` - returns a vector with the number of rows as the
    first element, and the number of columns as the second element (the
    **`dim`**ensions of the object)
-   `nrow(interviews)` - returns the number of rows
-   `ncol(interviews)` - returns the number of columns

Content:

-   `head(interviews)` - shows the first 6 rows
-   `tail(interviews)` - shows the last 6 rows

Names:

-   `names(interviews)` - returns the column names (synonym of
    `colnames()` for `data.frame` objects)

Summary:

-   `str(interviews)` - structure of the object and information about
    the class, length and content of each column
-   `summary(interviews)` - summary statistics for each column
-   `glimpse(interviews)` - returns the number of columns and rows of
    the tibble, the names and class of each column, and previews as many
    values will fit on the screen. Unlike the other inspecting functions
    listed above, `glimpse()` is not a `base R` function so you need to
    have the `tidyverse` package loaded to be able to execute it.

Note: most of these functions are "generic." They can be used on other
types of objects besides data frames or tibbles.

### Using functions

We can view the first few rows with the `head()` function, and the last
few rows with the `tail()` function:

```{r head}
head(interviews)
tail(interviews)
```

We used these functions with just one argument, the object `interviews`,
and we didn't give the argument a name. In `R`, a function's arguments
come in a particular order, and if you put them in the correct order,
you don't need to name them. In this case, the name of the argument is
`x`, so we can name it if we want, but since we know it's the first
argument, we don't need to.

Some arguments are optional. For example, the `n` argument in `head()`
specifies the number of rows to print. It defaults to 6, but we can
override that by specifying a different number:

```{r head-n}
head(interviews, n = 10)
```

If we order them correctly, we don't have to name either:

```{r head-n-noname}
head(interviews, 10)
```

Additionally, if we name them, we can put them in any order we want:

```{r head-reverse}
head(n = 10, x = interviews)
```

Generally, it's good practice to start with the required arguments, like
the data.frame whose rows you want to see, and then to name the optional
arguments. If you are ever unsure, it never hurts to explicitly name an
argument.

### Aside: Getting Help

To learn more about a function, you can type a `?` in front of the name
of the function, which will bring up the official documentation for that
function:

```{r, head-help}
?head
```

Function documentation is written by the authors of the functions, so
they can vary pretty widely in their style and readability. The first
section, **Description**, gives you a concise description of what the
function does, but it may not always be enough. The **Arguments**
section defines all the arguments for the function and is usually worth
reading thoroughly. Finally, the **Examples** section at the end will
often have some helpful examples that you can run to get a sense of what
the function is doing.

Another great source of information is **package vignettes**. Many
packages have vignettes, which are like tutorials that introduce the
package, specific functions, or general methods. You can run
`vignette(package = "package_name")` to see a list of vignettes in that
package. Once you have a name, you can run
`vignette("vignette_name", "package_name")` to view that vignette. You
can also use a web browser to go to
`https://cran.r-project.org/web/packages/package_name/vignettes/` where
you will find a list of links to each vignette. Some packages will have
their own websites, which often have nicely formatted vignettes and
tutorials.

Finally, learning to search for help is probably the most useful skill
for any `R` user. The key skill is figuring out what you should actually
search for. It's often a good idea to start your search with `R` or
`R programming`. If you have the name of a package you want to use,
start with `R package_name`.

Let's investigate `str` a bit more.

```{r}
str(interviews)
```

We get quite a bit of useful information here. First, we are told that
we have a data.frame of `r nrow(interviews)` observations, or rows, and
`r ncol(interviews)` variables, or columns.

Next, we get a bit of information on each variable, including its type
(`int` or `chr`) and a quick peek at the first 10 values. You might ask
why there is a `$` in front of each variable. This is because the `$` is
an operator that allows us to select individual columns from a
data.frame.

The `$` operator also allows you to use tab-completion to quickly select
which variable you want from a given data.frame. For example, to get the
`village` variable, we can type `interviews$` and then hit
<kbd>Tab</kbd>. We get a list of the variables that we can move through
with up and down arrow keys. Hit <kbd>Enter</kbd> when you reach
`village`, which should finish this code:

```{r dollar-subsetting}
interviews$village

```

## Vectors: the building block of data

You might have noticed that our last result looked different from when
we printed out the `interviews` data.frame itself. That's because it is
not a data.frame, it is a **vector**. A vector is a 1-dimensional series
of values, in this case a vector of characters representing the village
name.

Data.frames are made up of vectors; each column in a data.frame is a
vector. Vectors are the basic building blocks of all data in `R`.
Basically, everything in `R` is a vector, a bunch of vectors stitched
together in some way, or a function. Understanding how vectors work is
crucial to understanding how `R` treats data, so we will spend some time
learning about them.

There are 4 main types of vectors (also known as *atomic vectors*):

1.  `"character"` for strings of characters, like our `village` or
    `respondent_wall_type` columns. Each entry in a character vector is
    wrapped in quotes. In other programming languages, this type of data
    may be referred to as "strings".

2.  `"integer"` for integers. All the numeric values in `interviews` are
    integers. You may sometimes see integers represented like `2L` or
    `20L`. The `L` indicates to `R` that it is an integer, instead of the
    next data type, `"numeric"`.

3.  `"numeric"`, aka `"double"`, vectors can contain numbers including
    decimals. Other languages may refer to these as "float" or "floating
    point" numbers.

4.  `"logical"` for `TRUE` and `FALSE`, which can also be represented as
    `T` and `F`. In other contexts, these may be referred to as
    "Boolean" data.

Vectors can only be of a **single type**. Since each column in a
data.frame is a vector, this means an accidental character following a
number, like `29,` can change the type of the whole vector. Mixing up
vector types is one of the most common mistakes in `R`, and it can be
tricky to figure out. It's often very useful to check the types of
vectors.

To create a vector from scratch, we can use the `c()` function, putting
values inside, separated by commas.

```{r c-vector}
c(1, 2, 5, 12, 4)
```

As you can see, those values get printed out in the console, just like
with `interviews$village`. To store this vector so we can continue to
work with it, we need to assign it to an object.

```{r assign-vector}
num <- c(1, 2, 5, 12, 4)
```

You can check what kind of object `num` is with the `class()` function.

```{r}
class(num)
```

We see that `num` is a `numeric` vector.

Let's try making a character vector:

```{r}
char <- c("apple", "pear", "grape")
class(char)
```

Remember that each entry, like `"apple"`, needs to be surrounded by
quotes, and entries are separated with commas. If you do something like
`"apple, pear, grape"`, you will have only a single entry containing
that whole string.

Finally, let's make a logical vector:

```{r}
logi <- c(TRUE, FALSE, TRUE, TRUE)
class(logi)
```

:::::: challenge
## Challenge 1: Coercion

Since vectors can only hold one type of data, something has to be done
when we try to combine different types of data into one vector.

1.  What type will each of these vectors be? Try to guess without
    running any code at first, then run the code and use `class()` to
    verify your answers.

```{r coercion-challenge}
num_logi <- c(1, 4, 6, TRUE)
num_char <- c(1, 3, "10", 6)
char_logi <- c("a", "b", TRUE)


tricky <- c("a", "b", "1", FALSE)
```

::: solution
```{r coercion-challenge-answer}
class(num_logi)
class(num_char)
class(char_logi)
class(tricky)
```

`R` will automatically convert values in a vector so that they are all the
same type, a process called **coercion**.
:::

2.  How many values in `combined_logical` are `"TRUE"` (as a character)?

```{r combined-logical-challenge}
combined_logical <- c(num_logi, char_logi)
```

::: solution
```{r combined-logical-challenge-answer}
combined_logical
class(combined_logical)
```

Only one value is `"TRUE"`. Coercion happens when each vector is
created, so the `TRUE` in `num_logi` becomes a `1`, while the `TRUE` in
`char_logi` becomes `"TRUE"`. When these two vectors are combined, R
doesn't remember that the `1` in `num_logi` used to be a `TRUE`, it will
just coerce the `1` to `"1"`.
:::

3.  Now that you've seen a few examples of coercion, you might have
    started to see that there are some rules about how types get
    converted. There is a hierarchy to coercion. Can you draw a diagram
    that represents the hierarchy of what types get converted to other
    types?

::: solution
logical → integer → numeric → character

Logical vectors can only take on two values: `TRUE` or `FALSE`. Integer
vectors can only contain integers, so `TRUE` and `FALSE` can be coerced
to `1` and `0`. Numeric vectors can contain numbers with decimals, so
integers can be coerced from, say, `6` to `6.0` (though R will still
display a numeric `6` as `6`.). Finally, any string of characters can be
represented as a character vector, so any of the other types can be
coerced to a character vector.
:::
::::::

Coercion is not something you will often do intentionally; rather, when
combining vectors or reading data into `R`, a stray character that you
missed may change an entire numeric vector into a character vector. It
is a good idea to check the `class()` of your results frequently,
particularly if you are running into confusing error messages.

## Missing data

One of the great things about `R` is how it handles missing data, which
can be tricky in other programming languages. `R` represents missing data
as `NA`, without quotes, in vectors of any type. Let's make a numeric
vector with an `NA` value:

```{r na-vec}
weights <- c(25, 34, 12, NA, 42)
```

`R` doesn't make assumptions about how you want to handle missing data, so
if we pass this vector to a numeric function like `min()`, it won't know
what to do, so it returns `NA`:

```{r func-na}
min(weights)
```

This is a very good thing, since we won't accidentally forget to
consider our missing data. If we decide to exclude our missing values,
many basic math functions have an argument to **`r`** **`e`** **`m`**ove them:

```{r func-na-rm}
min(weights, na.rm = TRUE)
```

## Building with vectors

We have now seen vectors in a few different forms: as columns in a
data.frame and as single vectors. However, they can be manipulated into
lots of other shapes and forms. Some other common forms are:

-   matrices
    -   2-dimensional numeric representations
-   arrays
    -   many-dimensional numeric
-   lists
    -   lists are very flexible ways to store vectors
    -   a list can contain vectors of many different types and lengths
    -   an entry in a list can be another list, so lists can get deeply
        nested
    -   a data.frame is a type of list where each column is an
        individual vector and each vector has to be the same length,
        since a data.frame has an entry in every column for each row
-   factors
    -   a way to represent categorical data
    -   factors can be ordered or unordered
    -   they often *look* like character vectors, but behave differently
    -   under the hood, they are integers with character labels, called
        **levels**, for each integer

## Factors

`R` has a special data class, called `factor`, to deal with categorical data
that you may encounter when creating plots or doing statistical
analyses. Factors are very useful and actually contribute to making `R`
particularly well suited to working with data. So we are going to spend
a little time introducing them.

Factors represent categorical data. They are stored as integers
associated with labels and they can be ordered (ordinal) or unordered
(nominal). Factors create a structured relation between the different
levels (values) of a categorical variable, such as days of the week or
responses to a question in a survey. This can make it easier to see how
one element relates to the other elements in a column. While factors
look (and often behave) like character vectors, they are actually
treated as integer vectors by `R`. So you need to be very careful when
treating them as strings.

Once created, factors can only contain a pre-defined set of values,
known as *levels*. By default, `R` always sorts levels in alphabetical
order. For instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
respondent_floor_type <- factor(c("earth", "cement", "cement", "earth"))
```

`R` will assign `1` to the level `"cement"` and `2` to the level `"earth"`
(because `c` comes before `e`, even though the first element in this
vector is `"earth"`). You can see this by using the function `levels()`
and you can find the number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(respondent_floor_type)

nlevels(respondent_floor_type)
```

Sometimes, the order of the factors does not matter. Other times you
might want to specify the order because it is meaningful (e.g., `low`,
`medium`, `high`). It may improve your visualization, or it may be
required by a particular type of analysis. Here, one way to reorder our
levels in the `respondent_floor_type` vector would be:

```{r, results=TRUE, purl=FALSE}
respondent_floor_type # current order

respondent_floor_type <- factor(respondent_floor_type, 
                                levels = c("earth", "cement"))

respondent_floor_type # after re-ordering
```

In `R`'s memory, these factors are represented by integers (1, 2), but are
more informative than integers because factors are self describing:
`"cement"`, `"earth"` is more descriptive than `1`, and `2`. Which one
is "earth"? You wouldn't be able to tell just from the integer data.
Factors, on the other hand, have this information built in. It is
particularly helpful when there are many levels. It also makes renaming
levels easier. Let's say we made a mistake and need to recode `cement`
to `brick`. We can do this using the `fct_recode()` function from the
**`forcats`** package (included in the **`tidyverse`**) which provides
some extra tools to work with factors.

```{r, results=TRUE, purl=FALSE}
levels(respondent_floor_type)

respondent_floor_type <- fct_recode(respondent_floor_type, brick = "cement")

## as an alternative, we could change the "cement" level directly using the
## levels() function, but we have to remember that "cement" is the second level
# levels(respondent_floor_type)[2] <- "brick"

levels(respondent_floor_type)

respondent_floor_type
```

So far, your factor is unordered, like a nominal variable. `R` does not
know the difference between a nominal and an ordinal variable. You make
your factor an ordered factor by using the `ordered=TRUE` option inside
your factor function. Note how the reported levels changed from the
unordered factor above to the ordered version below. Ordered levels use
the less than sign `<` to denote level ranking.

```{r, results=TRUE, purl=FALSE}
respondent_floor_type_ordered <- factor(respondent_floor_type, 
                                        ordered = TRUE)

respondent_floor_type_ordered # after setting as ordered factor
```

### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(respondent_floor_type)
```

Converting factors where the levels appear as numbers (such as
concentration levels, or years) to a numeric vector is a little
trickier. The `as.numeric()` function returns the index values of the
factor, not its levels, so it will result in an entirely new (and
unwanted in this case) set of numbers. One method to avoid this is to
convert factors to characters, and then to numbers. Another method is to
use the `levels()` function. Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))

as.numeric(year_fct)                     # Wrong! And there is no warning...

as.numeric(as.character(year_fct))       # Works...

as.numeric(levels(year_fct))[year_fct]   # The recommended way.
```

Notice that in the recommended `levels()` approach, three important
steps occur:

-   We obtain all the factor levels using `levels(year_fct)`
-   We convert these levels to numeric values using
    `as.numeric(levels(year_fct))`
-   We then access these numeric values using the underlying integers of
    the vector `year_fct` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function
to get a quick glance at the number of observations represented by each
factor level. Let's extract the `memb_assoc` column from our data frame,
convert it into a factor, and use it to look at the number of interview
respondents who were or were not members of an irrigation association:

```{r factor-plot-default-order, fig.alt="Yes/no bar graph showing number of individuals who are members of irrigation association", purl=TRUE}
## create a vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc

## convert it into a factor
memb_assoc <- as.factor(memb_assoc)

## let's see what it looks like
memb_assoc

## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```

Looking at the plot compared to the output of the vector, we can see
that in addition to `"no"s` and `"yes"s`, there are some respondents for
whom the information about whether they were part of an irrigation
association hasn't been recorded, and encoded as missing data. These
respondents do not appear on the plot. Let's encode them differently so
they can be counted and visualized in our plot.

```{r factor-plot-reorder, fig.alt="Bar plot of association membership, showing missing responses.", purl=TRUE}
## Let's recreate the vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc

## replace the missing data with "undetermined"
memb_assoc[is.na(memb_assoc)] <- "undetermined"

## convert it into a factor
memb_assoc <- as.factor(memb_assoc)

## let's see what it looks like
memb_assoc

## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```

:::: challenge
## Exercise

-   Rename the levels of the factor to have the first letter in
    uppercase: `No`, `Undetermined`, and `Yes`.

-   Now that we have renamed the factor level to `Undetermined`, can you
    recreate the barplot such that `Undetermined` is last (after `Yes`)?

::: solution
## Solution

```{r factor-plot-exercise, fig.alt="bar graph showing number of individuals who are members of irrigation association, including undetermined option"}
## Rename levels.
memb_assoc <- fct_recode(memb_assoc, No = "no",
                         Undetermined = "undetermined", Yes = "yes")
## Reorder levels. Note we need to use the new level names.
memb_assoc <- factor(memb_assoc, levels = c("No", "Yes", "Undetermined"))
plot(memb_assoc)
```
:::
::::

## Formatting Dates

One of the most common issues that new (and experienced!) `R` users have
is converting date and time information into a variable that is
appropriate and usable during analyses. A best practice for dealing with
date data is to ensure that each component of your date is available as
a separate variable. In our dataset, we have a column `interview_date`
which contains information about the year, month, and day that the
interview was conducted. Let's convert those dates into three separate
columns.

```{r, eval=FALSE, purl=FALSE}
str(interviews)
```

We are going to use the package **`lubridate`**, which is included in
the **`tidyverse`** installation and should be loaded by default.

The `lubridate` function `ymd()` takes a vector representing year, month,
and day, and converts it to a `Date` vector. `Date` is a class of data
recognized by `R` as being a date and can be manipulated as such. The
argument that the function requires is flexible, but, as a best
practice, is a character vector formatted as `YYYY-MM-DD`.

To learn more about **`lubridate`** after this workshop, you may want to
check out this handy [`lubridate` cheatsheet](https://rstudio.github.io/cheatsheets/html/lubridate.html?_gl=1*ud3j0e*_ga*MTgwMjI0NTA1LjE3NzMwMjUwNzQ.*_ga_2C0WZ1JHG0*czE3NzMyODY1ODUkbzMkZzAkdDE3NzMyODY1ODUkajYwJGwwJGgw).

Let's extract our `interview_date` column and inspect the structure:

```{r, purl=FALSE}
dates <- interviews$interview_date
str(dates)
```

When we imported the data in `R`, `read_csv()` recognized that this column
contained date information. We can now use the `day()`, `month()` and
`year()` functions to extract this information from the date, and create
new columns in our data frame to store it:

```{r, purl=FALSE}
interviews$day <- day(dates)
interviews$month <- month(dates)
interviews$year <- year(dates)
interviews
```

Notice the three new columns at the end of our data frame.

In our example above, the `interview_date` column was read in correctly
as a `Date` variable but generally that is not the case. Date columns
are often read in as `character` variables and one can use the
`as_date()` function to convert them to the appropriate
`Date/POSIXct`format.

Let's say we have a vector of dates in character format:

```{r, purl=FALSE}
char_dates <- c("7/31/2012", "8/9/2014", "4/30/2016")
str(char_dates)
```

We can convert this vector to dates as :

```{r, purl=FALSE}
as_date(char_dates, format = "%m/%d/%Y")
```

Argument `format` tells the function the order to parse the characters
and identify the month, day and year. The format above is the equivalent
of `mm/dd/yyyy`. A wrong format can lead to parsing errors or incorrect
results.

For example, observe what happens when we use a lower case `y` instead of
upper case `Y` for the year.

```{r, purl=FALSE}
as_date(char_dates, format = "%m/%d/%y")
```

Here, the `%y` part of the format stands for a two-digit year instead of
a four-digit year, and this leads to parsing errors.

Or in the following example, observe what happens when the month and day
elements of the format are switched.

```{r, purl=FALSE}
as_date(char_dates, format = "%d/%m/%y")
```

Since there is no month numbered 30 or 31, the first and third dates
cannot be parsed.

We can also use functions `ymd()`, `mdy()` or `dmy()` to convert
character variables to date.

```{r, purl=FALSE}
mdy(char_dates)
```

::: keypoints
-   Use `read_csv` to read tabular data in `R`.
-   Use `factors` to represent categorical data in `R`.
:::
